In the ever-evolving landscape of C++, mastering CMake is indispensable for any developer striving to write efficient, maintainable, and scalable code. Modern CMake for C++ by Rafał Świdziński serves as a beacon, guiding both novices and seasoned programmers through the intricacies of CMake.

This book is not just a manual; it is a journey. It starts with the basics, ensuring that even those new to CMake can grasp its concepts. As the chapters progress, readers are equipped with advanced techniques, empowering them to harness the full potential of CMake.
What sets this book apart is its pragmatic approach. Real-world examples and best practices are interwoven throughout the text, ensuring that readers not only understand the concepts but also know how to apply them effectively in their projects.

By the end of this book, readers will not only have a deep understanding of CMake but also a newfound confidence in their ability to navigate the complexities of C++ development. They will be armed with the knowledge and skills needed to write cleaner, more efficient code, setting them on a path to becoming proficient CMake developers.

Modern CMake for C++ is not just a book; it is a tool that will empower its readers to elevate their C++ development skills to new heights. Whether you’re a beginner or an expert, this book will help you unlock the full potential of CMake, making your code more robust, maintainable, and scalable.

\begin{center}
\textit{Alexander Kushnir}

\textit{Principal Software Engineer, Biosense Webster}
\end{center}

