Programming is as much an art as it is a deeply technical process, and as we know all too well, it’s very difficult. Therefore, we should be looking to optimize this process as much as possible. There aren’t too many instances where we can just flip a switch and get better outcomes, but using Integrated Development Environments (IDEs) is definitely one of those rare cases.

If you haven’t worked with a proper IDE before (or you believe that a text processor like Emacs or Vim is the best you can hope for), this chapter is for you. If you’re a seasoned professional and are already familiar with the subject, you can use this chapter as a quick overview of the current top choices and maybe consider a switch or, even better, get clear confirmation that your current tool is the best.

With an emphasis on accessibility for those new to the field, this chapter provides a gentle introduction to the critical choice of an IDE. We will cover why you’d want one and how to pick one that best suits your needs. Sure, there are many choices out there, but as usual, some are just better than others. Unfortunately, this isn’t a universal, one-size-fits-all kind of deal. Many factors contribute toward the productivity levels that you might get if you choose right. We’ll discuss a few considerations that might be important if your work is in an organization of a certain size, ensuring that you grasp the nuances without becoming mired in complexity. This will be followed by a quick introduction to toolchains, where we’ll discuss available choices.

We will then highlight the distinctive qualities of several popular IDEs, such as the sophisticated CLion, the adaptable nature of Visual Studio Code, and then the powerhouse that is the Visual Studio IDE. Each section is tailored to showcase the strengths and advanced features that these In this chapter, we’re going to cover the following main topics:

\begin{itemize}
\item
Getting to know IDEs

\item
Starting with the CLion IDE

\item
Starting with Visual Studio Code

\item
Starting with the Visual Studio IDE
\end{itemize}




